Climatic warming strengthens a positive feedback between alpine shrubs and fire

### 2016-03-09 manuscript/manuscript.Rnw
More minor edits

\begin{abstract}
Climate change is expected to increase fire activity and exacerbate woody plant encroachment in arctic and alpine grasslands. What is unknown is whether increases in woody plants and fire will interact to exacerbate changes in structure, function and composition of these ecosystems. Here we use field surveys and experimental manipulations to examine warming and fire effects on recruitment, growth and survival of seedlings of evergreen obligate seeding alpine shrubs. We find that fire substantially increased shrub seedling establishment (up to \Sexpr{density_ratios(greaus_seedling_density_model, 'Grevillea')}) and that warming doubled tall shrub seedling growth rates and can increase survival. Warming also had weak or no effect on shrub recruitment, rates of grass post-fire gap infilling or grass competitive effects on shrub seedling growth and survival. These findings indicate that rising temperatures coupled with more frequent or severe fires will likely exacerbate increases in tall evergreen shrubs by increasing recruitment, doubling growth rates, and potentially increasing seedling survival.
\end{abstract}

@@

In alpine and arctic ecosystems, warming experiments and long-term monitoring have documented significant increases in the growth and cover of woody species\cite{Sturm:2001vd,Walker:2006ht,MyersSmith:2011fp}. The frequency and extent of wildfires have also increased in these environments over recent decades, a trend expected to continue\cite{Westerling:2006ve, Flannigan:2009kl, Qiu:2009th, Bradstock:2014jp}. In alpine and tundra ecosystems, current evidence from long-term and experimental studies indicate that shrub recruitment and encroachment is highest in disturbed areas\cite{Williams:1987ua, Batllori:2009fe, Frost:2013tm}, that climatic warming is likely to improve woody species growth rates\cite{Arft:1999iz, Elmendorf:2012ba, MyersSmith:2011fp}, and that shrubs are the most flammable vegetation component of these ecosystems\cite{Williams:2006tl, Higuera:2008dw, Higuera:2009ga}. The combination of this evidence indicates that more frequent or severe fire will increase recruitment opportunities for woody species, which when coupled with higher growth and survival rates caused by rising temperatures, will exacerbate shrub expansion\cite{Racine:2004tn,Lantz:2013fh} and ultimately increase the likelihood of more fires\cite{Higuera:2008dw, Higuera:2009ga,MyersSmith:2011fp}. In effect, warming will strengthen an existing feedback loop that not only has the potential to cause rapid changes in the composition and structure of alpine and arctic vegetation, but also have serious social, biodiversity and carbon sequestration consequences\cite{Mack:2011ew}.

While paleoecological studies have indicated that such a feedback may exist\cite{Higuera:2008dw, Higuera:2009ga}, there is a paucity of information on multiple mechanisms that may strengthen, mitigate or break this feedback loop (Fig. \ref{fig1}a). For example, we have little information on whether fire will stimulate shrub seedling recruitment in alpine or tundra ecosystems, nor do we know how rates of seedling growth and survival will be affected under warmer post-fire conditions. Furthermore, we do not know how tussock grasslands (the other dominant community in these ecosystems) will respond to warmer post-fire conditions, and whether grasses have competitive, facilitative or no effect on shrub seedling demographic rates, or whether such effects are altered by warmer post-fire conditions.

@@

\subsection{Drivers of shrub seedling establishment.}
We first investigated how altitude, topographic wetness (TWI), adult density, fire and, if burnt, fire severity, influenced the density of \textit{Grevillea} and \textit{Asterolasia} seedlings (the two dominant shrubs of open heathland). In 2012-13, we estimated seedling density for both species across 40 alpine sites and found that the abundance of \textit{Grevillea} (Fig. \ref{fig2}) and \textit{Asterolasia} (Supplementary Fig. S1) seedlings were strongly influenced by the occurrence of fire. Sites burnt by fire in 2003 had seedling densities between \Sexpr{density_ratios(asttry_seedling_density_model,'Asterolasia')} and \Sexpr{density_ratios(greaus_seedling_density_model,'Grevillea')} times higher than unburnt sites. In \textit{Grevillea}, mean seedling densities were \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', TRUE)}/m$^2$ at burnt sites and \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', FALSE)}/m$^2$ at unburnt sites. \textit{Asterolasia} had similar mean densities: \Sexpr{density_means(asttry_seedling_density_model,'Asterolasia', TRUE)} and \Sexpr{density_means(asttry_seedling_density_model, 'Asterolasia', FALSE)} seedlings/m$^2$ at burnt and unburnt sites, respectively. For both species, seedling density was also higher in severely burnt sites. As expected, pre-fire adult density positively influenced \textit{Grevillea} seedling density, with sites containing higher abundances of pre-fire \textit{Grevillea} adults containing more seedlings. For both shrub species, we detected no significant change in seedling density along a 200 m altitudinal range (equivalent to a \Sexpr{round(0.77 * (200/100),1)}\textdegree C change in mean temperature\cite{Brown:1989vx}). The field warming experiment (see below) indicates that this lack of elevation response in \textit{Grevillea} is due to both rates of recruitment (Supplementary Fig. S2) and mortality (see below) largely being insensitive to a \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C change in temperature. We detected no significant effect of topographic wetness for either species.

@@

Climate change is expected to alter biotic interactions because their strength and direction depend strongly on climatic conditions, particularly in alpine and arctic ecosystems\cite{Callaway:2002vb, Klanderud:2005uw}. Here, we assess the interactive effects of warming and inter-tussock gap size (i.e. the proximity of shrub seedlings to grass tussocks) on the growth and survival of \textit{Grevillea} seedlings (Fig. \ref{fig5}).  As with the seedlings growing with no nearby \textit{poa tussock}, we detected a strong positive effect of warming on growth rates and a marginal, yet not significant, decrease in mortality. We also found no significant effect of inter-tussock gap radius on growth or mortality rates. However, for growth, we did detect a marginally significant interaction between warming and gap size, such that in warmed plots growth rates slightly increased with gap size. This suggests that warmer conditions may increase competition between shrubs and grasses, a finding consistent with the abiotic stress hypothesis\cite{Callaway:2002vb}. However, this effect was small and unlikely to have significant biological consequence on shrub growth rates.

@@

Despite having little impact on shrub seedling growth and mortality rates, tussock grass may still limit shrub recruitment, and thus, shrub expansion, by infilling post-fire bare ground gaps at a faster rate under warmer conditions. Using five years of post-fire inter-tussock gap size changes in warmed and unwarmed plots we found that gaps were being infilled by grasses (Fig. \ref{fig6}). However, the rate at which this occurred was very slow, with a 10 cm radius gap decreasing by approximately \Sexpr{gap_size(gap_dynamic_model, 'otc')} cm over a ten year period. We also detected no significant effect of a \Sexpr{micro_clim(overall_microclimate_summary,'CAmbient_Temp','mean','grand_mean')}\textdegree C temperature rise on the rate of grass infilling.

@@

We used 40 long-term open heathland sites\cite{Camac:2013eb} established after the 2003 fires consisting of 30 burnt sites and 10 sites thought to be unburnt for over 70 years. In the summer of 2012-13, at each site, seedling density/m$^2$ was estimated using 40 1 m$^2$ quadrats, evenly distributed in groups of 10 along four 50 m transects, with 10 m between transect lines, subsampling an area of 2000 m$^2$. Within plots we recorded the number and maximum height of \textit{Grevillea} and \textit{Asterolasia} seedlings. For unburnt sites we counted the number of mature \textit{Grevillea} plants (\textgreater 0.5 m$^2$) within 5 m either side of each transect. In burnt sites, this required counting the number of skeletons (there were no living adults at any burnt site) that still persist post-fire. We were unable estimate numbers of adult \textit{Asterolasia} because this species does not have a persistent woody skeleton post-fire. Site level data, elevation and Topographic Wetness Index (TWI), were obtained from a 30 m resolution digital elevation model. Lastly, for burnt sites, fire severity was estimated using minimum twig diameters, collected immediately after the 2003 fires.

@@

In total we built 11 hierarchical models to examine factors influencing tussock grass post-fire recovery and shrub seedling establishment, growth and mortality. For each model (Supplementary Methods) we used Bayesian inference and fitted models in R \Sexpr{paste(version$major, version$minor, sep='.')} using package Rstan \Sexpr{packageVersion("rstan")}. Data and source code for reproducing analysis and figures is available at: \url{https://github.com/jscamac/Alpine_Shrub_Experiment}. Additional information of experimental design and analysis can be found in Supplementary methods.
\end{methods}

>>>>

\begin{abstract}
Climate change is expected to increase fire activity and exacerbate woody plant encroachment in arctic and alpine landscapes. What is unknown is whether increases in woody plants and fire will interact to exacerbate changes in structure, function and composition of these ecosystems. Here we use field surveys and experimental manipulations to examine warming and fire effects on recruitment, growth and survival of seedlings of evergreen obligate seeding alpine shrubs. We find that fire substantially increased shrub seedling establishment (up to \Sexpr{density_ratios(greaus_seedling_density_model, 'Grevillea')}) and that warming doubled tall shrub seedling growth rates and can increase survival. Warming also had weak or no effect on shrub recruitment, rates of grass post-fire gap infilling or grass competitive effects on shrub seedling growth and survival. These findings indicate that rising temperatures coupled with more frequent or severe fires will likely exacerbate increases in tall evergreen shrubs by increasing recruitment, doubling growth rates, and potentially increasing seedling survival.
\end{abstract}

@@

In alpine and arctic ecosystems, warming experiments and long-term monitoring have documented significant increases in the growth and cover of woody species\cite{Sturm:2001vd,Walker:2006ht,MyersSmith:2011fp}. The frequency and extent of wildfires have also increased in these environments over recent decades, a trend expected to continue\cite{Westerling:2006ve, Flannigan:2009kl, Qiu:2009th, Bradstock:2014jp}. In alpine and tundra ecosystems, current evidence from long-term and experimental studies indicate that shrub recruitment and encroachment is highest in disturbed areas\cite{Williams:1987ua, Batllori:2009fe, Frost:2013tm}, that climatic warming is likely to improve woody species growth rates\cite{Arft:1999iz, Elmendorf:2012ba, MyersSmith:2011fp}, and that shrubs are the most flammable vegetation component of these ecosystems\cite{Williams:2006tl, Higuera:2008dw, Higuera:2009ga}. The combination of this evidence indicates that more frequent or severe fire may increase recruitment opportunities for woody species, which when coupled with higher growth and survival rates caused by rising temperatures, will exacerbate shrub expansion\cite{Racine:2004tn,Lantz:2013fh} and ultimately increase the likelihood of more fires\cite{Higuera:2008dw, Higuera:2009ga,MyersSmith:2011fp}. In effect, warming could strengthen an existing feedback loop that not only has the potential to cause rapid changes in the composition and structure of alpine and arctic vegetation, but also have serious social, biodiversity and carbon sequestration consequences\cite{Mack:2011ew}.

While paleoecological studies have indicated that such a feedback may exist\cite{Higuera:2008dw, Higuera:2009ga}, there is a paucity of information on multiple mechanisms that may strengthen, mitigate or break this feedback loop (Fig. \ref{fig1}a). For example, we have little information on whether fire will stimulate shrub seedling recruitment in alpine or tundra ecosystems, nor do we know how rates of seedling growth and survival will be affected under warmer post-fire conditions. Furthermore, we do not know how tussock grasslands (the other dominant community in these ecosystems) will respond to warmer post-fire conditions, and whether grasses have competitive, facilitative or no effect on shrub seedling vital rates, or whether such effects are altered by warmer post-fire conditions.

@@

\subsection{Drivers of shrub seedling establishment.}
We first investigated how altitude, topographic wetness (TWI), adult density, fire and, if burnt, fire severity, influenced the density of \textit{Grevillea} and \textit{Asterolasia} seedlings (the two dominant shrubs of open heathland). In 2011-12, we estimated seedling density for both species across 40 alpine sites and found that the abundance of \textit{Grevillea} (Fig. \ref{fig2}) and \textit{Asterolasia} (Supplementary Fig. S1) seedlings were strongly influenced by the occurrence of fire. Sites burnt by fire in 2003 had seedling densities between \Sexpr{density_ratios(asttry_seedling_density_model,'Asterolasia')} and \Sexpr{density_ratios(greaus_seedling_density_model,'Grevillea')} times higher than unburnt sites. In \textit{Grevillea}, mean seedling densities were \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', TRUE)}/m$^2$ at burnt sites and \Sexpr{density_means(greaus_seedling_density_model,'Grevillea', FALSE)}/m$^2$ at unburnt sites. \textit{Asterolasia} had similar mean densities: \Sexpr{density_means(asttry_seedling_density_model,'Asterolasia', TRUE)} and \Sexpr{density_means(asttry_seedling_density_model, 'Asterolasia', FALSE)} seedlings/m$^2$ at burnt and unburnt sites, respectively. For both species, seedling density was also higher in severely burnt sites. As expected, pre-fire adult density positively influenced \textit{Grevillea} seedling density, with sites containing higher abundances of pre-fire \textit{Grevillea} adults containing more seedlings. For both shrub species, we detected no significant change in seedling density along a \Sexpr{round(max(complete_field_survey_data$altitude_m)) - round(min(complete_field_survey_data$altitude_m))} m altitudinal range (equivalent to a \Sexpr{round(0.77 * (200/100),1)}\textdegree C change in mean temperature\cite{Brown:1989vx}). The field warming experiment (see below) indicates that this lack of elevation response in \textit{Grevillea} is due to both rates of recruitment (Supplementary Fig. S2) and mortality (see below) largely being insensitive to a \Sexpr{micro_clim(overall_microclimate_summary, 'CAmbient_Temp','mean','grand_mean')}\textdegree C change in temperature. We detected no significant effect of topographic wetness for either species.

@@

Climate change is expected to alter biotic interactions because their strength and direction depend strongly on climatic conditions, particularly in alpine and arctic ecosystems\cite{Callaway:2002vb, Klanderud:2005uw}. Here, we assess the interactive effects of warming and inter-tussock gap size (i.e. the proximity of shrub seedlings to grass tussocks) on the growth and survival of \textit{Grevillea} seedlings (Fig. \ref{fig5}).  As with the seedlings growing with no nearby \textit{Poa} tussock, we detected a strong positive effect of warming on growth rates and a marginal, yet not significant, decrease in mortality. We also found no significant effect of inter-tussock gap radius on growth or mortality rates. However, for growth, we did detect a marginally significant interaction between warming and gap size, such that in warmed plots growth rates slightly increased with gap size. This suggests that warmer conditions may increase competition between shrubs and grasses, a finding consistent with the abiotic stress hypothesis\cite{Callaway:2002vb}. However, this effect was small and unlikely to have significant biological consequence on shrub growth rates.

@@

Despite having little impact on shrub seedling growth and mortality rates, tussock grass may still limit shrub recruitment, and thus, shrub expansion, by infilling post-fire bare ground gaps (whether by vegatative growth or seedlings) at a faster rate under warmer conditions. Using five years of post-fire inter-tussock gap size changes in warmed and unwarmed plots we found that gaps were being infilled by grasses (Fig. \ref{fig6}). However, the rate at which this occurred was very slow, with a 10 cm radius gap predicted to decrease by approximately \Sexpr{gap_size(gap_dynamic_model, 'otc')} cm over a ten year period. We also detected no significant effect of a \Sexpr{micro_clim(overall_microclimate_summary,'CAmbient_Temp','mean','grand_mean')}\textdegree C temperature rise on the rate of grass infilling.

@@

We used 40 long-term open heathland sites\cite{Camac:2013eb} established after the 2003 fires consisting of 30 burnt sites and 10 sites thought to be unburnt for over 70 years. In the summer of 2011-12, at each site, seedling density/m$^2$ was estimated using 40 1 m$^2$ quadrats, evenly distributed in groups of 10 along four 50 m transects, with 10 m between transect lines, subsampling an area of 2000 m$^2$. Within plots we recorded the number and maximum height of \textit{Grevillea} and \textit{Asterolasia} seedlings. For unburnt sites we counted the number of mature \textit{Grevillea} plants (\textgreater 0.5 m$^2$) within 5 m either side of each transect. In burnt sites, this required counting the number of skeletons (there were no living adults at any burnt site) that still persist post-fire. We were unable estimate numbers of adult \textit{Asterolasia} because this species does not have a persistent woody skeleton post-fire. Site level data, elevation and Topographic Wetness Index (TWI), were obtained from a 30 m resolution digital elevation model. Lastly, for burnt sites, fire severity was estimated using minimum twig diameters, collected immediately after the 2003 fires.

@@

In total we built 11 hierarchical models to examine factors influencing tussock grass post-fire recovery and shrub seedling establishment, growth and mortality. For each model (Supplementary Methods) we used Bayesian inference and fitted models in R \Sexpr{paste(version$major, version$minor, sep='.')} using package Rstan \Sexpr{packageVersion("rstan")}. Data and source code for reproducing analysis and figures are available at: \url{https://github.com/jscamac/Alpine_Shrub_Experiment}. Additional information of experimental design and analysis can be found in Supplementary methods.
\end{methods}
